Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-06 summary.tex
major revision underway, update to include global correction in numerical sims

In accordance with the above studies, which show that the VSI   
applies to perturbations with small radial lengthscales, we retained
the radially-local approximation by using the `vertically-global
shearing box' (VGSB) formalism developed by \cite{mcnally14}, who 
showed the linear isothermal VSI studied in \cite{nelson13}, can
be captured by the VGSB. We added 
an energy source term in the VGSB equations which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This form of thermal relaxation was also 
employed in the global numerical simulations of \cite{nelson13}, and allows
the thermodynamic response of the disk to be varied in a controlled 
manner.

We mostly focused on vertically isothermal disks ($\Gamma=1$) because
it permits a largely analytical discussion. Our analysis is an
extension of the linear models considered in \cite{lubow93} by
including vertical shear, but in the limit of low-frequency
perturbations in  nearly-Keplerian disks. These approximations were
made for simplicity, but also motivated by previous work
\citep{nelson13}.

For isothermal perturbations ($\gamma=1$ or $\beta=0$) we explicitly
demonstrated instability 
associated with vertical shear, and that the VSI growth rate is
limited by the maximum vertical shear within the vertical domain of
interest. In the opposite limit of adiabatic perturbations ($\gamma>1$
and $\beta\to\infty$), we find VSI growth rates are dramatically
reduced on account of a stabilizing 
vertical entropy gradient. The VSI is not completely
surpressed because vertical shear dominates over buoyancy 
sufficiently close the disk midplane. However, perturbations grow
slowly and is constrained to regions immediately about the disk
midplane. Thus, the adiabatic VSI is likely unimportant in practice.

For non-isothermal ($\gamma\neq1$) perturbations with intermediate
$\beta$, the general dispersion relation
(Eq. \ref{relax_disp}) is complicated and requires a numerical
treatment. However, for the fundamental VSI, characterized by  $\delta
v_z\sim\mathrm{constant}$ near $z=0$, some simplification is
possible. In particular, we quantified the thermal relaxation
timescale $\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below
which small-scale perturbations ($\khat\gg 1$) may be regarded as isothermal
and the standard VSI operates. We believe this explicit expression for
$\beta_\mathrm{crit}$ is the most useful result from our
investigation.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. Numerical solutions are mostly consistent
with our analytical expectations, except for perturbations with small
radial wavenumber ($\khat\to0$) where the low-frequency approximation
is inappropriate.

We briefly studied vertically non-isothermal disks using our linear
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to thermal relaxation. When a
stable stratification is introduced, the reduction of VSI growth
rates by a finite thermal relaxation timescale is qualitatively similar to
that observed for vertically isothermal disks.

@@

We applied our linear results to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae (MMSN),
using disk models described in \cite{chiang10}. We find the MMSN is
unstable to the VSI at $10$s of AU, with characteristic
growth times of $10$s of orbits. However, we note the strong
temperature dependence of the physical thermal relaxation timescale
in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
radial scale). Thus, increasing the disk temperature can signifcantly 
shorten the typical growth times for the VSI. This is also attributed
to a larger value of $\beta_\mathrm{crit}$, since increasing $T$ also
increases the aspect-ratio $\epsilon$.

@@

>>>>

In accordance with the above studies, which show that the VSI   
applies to perturbations with small radial lengthscales, we retained
the radially-local approximation by using the `vertically-global
shearing box' (VGSB) formalism developed by \cite{mcnally14}, who 
showed the linear VSI studied by \cite{nelson13} in a locally
isothermal disk can be captured by the VGSB.

% The presence of a
% vertical shear, but 
% is not consistent with the
% global

We added an energy source term in the VGSB equations which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This form of thermal relaxation was also 
employed in the global numerical simulations of \cite{nelson13}, and allows
the thermodynamic response of the disk to be controlled.

We mostly focused on vertically isothermal disks ($\Gamma=1$) because
it permits a largely analytical discussion. Our analysis is an
extension of the linear models considered in \cite{lubow93} by
adding vertical shear, but in the limit of low-frequency
perturbations in nearly-Keplerian disks. These approximations were
made for simplicity, but also motivated by previous work
\citep{nelson13}.

For isothermal perturbations ($\gamma=1$ or $\beta=0$) we explicitly
demonstrated instability  associated with vertical shear, and that the
VSI growth rate is limited by the maximum vertical shear within the
vertical domain of interest. The fundamental VSI mode is characterized
by one node in the density perturbation at the mid-plane
(i.e. anti-symmetric about $z=0$), which corresponds to a vertical
velocity perturbation  $\delta v_z\sim\mathrm{constant}$ near $z=0$.

% In the opposite limit of adiabatic perturbations ($\gamma>1$
% and $\beta\to\infty$), we find VSI growth rates are dramatically
% reduced on account of a stabilizing 
% vertical entropy gradient. The VSI is not completely
% surpressed because vertical shear dominates over buoyancy 
% sufficiently close the disk midplane. However, perturbations grow
% slowly and is constrained to regions immediately about the disk
% midplane. Thus, the adiabatic VSI is likely unimportant in practice.

For non-isothermal  perturbations with thermal relaxation ($\gamma>1$,
$\beta>0$) , the general dispersion relation (Eq. \ref{relax_disp}) is
complicated and requires a numerical treatment. However, for the
fundamental VSI some simplification is  possible. In particular, we quantified the critical thermal relaxation
timescale $\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below
which small-scale perturbations ($\khat\gg 1$) may be regarded as isothermal
and the fundamental VSI operates. We believe this explicit expression for
$\beta_\mathrm{crit}$ is the most useful result from our
investigation.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. Numerical solutions are mostly consistent
with our analytical expectations, except for perturbations with small
radial wavenumber ($\khat\to0$) where the low-frequency approximation
is inappropriate. The local model is also not valid for such modes.

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to
thermal relaxation. When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation timescale
is qualitatively similar to that observed for vertically isothermal
disks.

@@

We applied our linear results to assess the applicability of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae (MMSN),
using disk models described in \cite{chiang10}. We find the MMSN is
unstable to the fundamental VSI at $10$s of AU, with characteristic
growth times of $10$s of orbits. However, we note the strong
temperature dependence of the physical thermal relaxation timescale
in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
radial scale). Thus, increasing the disk temperature even slightly can signifcantly 
shorten the typical growth times for the VSI. This is also attributed
to a larger value of $\beta_\mathrm{crit}$, since increasing $T$ also
increases the aspect-ratio $\epsilon$.

@@

%need to compare kley sims
